---
title: "Homework 5"
output: pdf_document
pdf_document: default
html_document: default
---

# Homework 5
Student: Jama Brookes

This homework requires `wine.csv`, and the `tidyverse` and `Rtsne` packages. Install them if you haven't already!  
See the following link for how to add new packages to Binder: https://github.com/rjenki/BIOS512?tab=readme-ov-file#adding-packages-to-installr-later.   
**For readability and easier processing, please make each question part a different code chunk.**

```{r}
#exporting to rmd
#library(rmarkdown)
#convert_ipynb("./BIOS512_hw5_Brookes.ipynb", output = xfun::with_ext("./BIOS512_HW5_Brookes.ipynb", "Rmd"))
```

```{r}
library(tidyverse)
library(Rtsne)
```

## Question 1  

#### a) Import your data.  
```{r}
wine <- read.csv("./hw5_data/wine.csv")
```

#### b) Check out the columns present using one of R's data frame summary.
```{r}
str(wine)

#changing class to character

wine$class <- as.character(wine$class)

str(wine)

```

#### c) Get summary statistics on the numeric variables.  
```{r}
wine %>% select(where(is.numeric)) %>% 
  summary()
```


## Question 2

#### a) Scale and center your data  
*Hint:* Use a `mutate()` statement across all columns **except class** with `function(x) as.numeric(scale(x))`.
```{r}
wine_scaled <- wine %>% 
  mutate(across(where(is.numeric), ~as.numeric(scale(.))))

summary(wine_scaled)
```


#### b) Based on what you saw in the summary statistic table from the imported data, why would scaling and centering this data be helpful before we perform PCA?

It is helpful because the raw data is not centered around a common point. All the summary values
range to varying degrees above zero, and it is difficult to work with the data
without a standardized or centered scale. We have to scale and center so PCA can rotate around
a common center.


## Question 3

#### a) Perform PCA

```{r}
PCA_wine <- wine_scaled %>% select(where(is.numeric)) %>% 
                                  prcomp(); # runs PCA on d
PCA_wine
summary(PCA_wine)
```


#### b) How much of the total variance is explained by PC1? PC2? What function do we use to see that information?
```{r}
#PC1 Proportion of variance + PC2 proportion of variance
0.361+0.1921
```

Total Proportion of Variance explained by PC1 and PC2 is 0.5531.


#### c) Why are we doing PCA first?
Reduces noise and dimensionality in the data.
It also helps to capture the most variation in the data to create new axes.

#### d) What is the rotation matrix? Print it explicitly.  
*Hint:* Check the notes for a simple way to do this!
```{r}

print(PCA_wine$rotation) # PCA_wine$rotation is the rotation matrix

```


#### e) Plot PC1 vs. PC2, using the wine class as labels for coloring.  
*Hint:* You'll first need a data set with only PC1 and PC2, then add back the class variable from your scaled data set with a `mutate()` statement. Then, you can use `color = factor(class)` in your `ggplot` statement.
```{r}
PC1_PC2_only <- PCA_wine$x %>% as.data.frame() %>%  select(c("PC1", "PC2"))

PC1_PC2_only$class <-wine_scaled$class

#plotting PC1 vs PC2

library(ggplot2)
ggplot(PC1_PC2_only, aes(x = PC1, y = PC2, color = factor(class))) +
  geom_point() +
  labs(color = "Class") +
  theme_bw()
```

#### f) What do you see after plotting PC1 vs. PC2? What does this mean in context of wine classes?
The wine classes are clustered on the graph, but overlap some on the PC1 and PC2 axis. 
Class 2 shares some similarities with Class 1 and 2. However, there seems to be
distinct clustering on chemical profiles that define the different wine classes.

#### g) Give an example of data where PCA would fail. You can describe the data or do a simulation.  
*Hint:* Our notes have a few examples!
PCA would fail on a dataset with a non-linear space. 
A real life example of this would be calculating differences in a dataset of facial expressions.
This dataset would not be linear because facial expression patterns could not be maniuplated
on a linear space.
Example simulation:

```{r}
circular_data <- tibble(r = sample(c(2,5),size=650,replace=TRUE) + rnorm(650,0,0.5),
            theta = runif(650,0,2*pi)) %>%
    transmute(x=r*cos(theta),
              y=r*sin(theta));
ggplot(circular_data, aes(x,y)) + geom_point() + coord_fixed()
```


#### h) Explain the difference between vector space and manifold, and how these terms apply to what we did/will do with T-SNE.
A vector space is a set of variables along a field that follow a mathematical structure that assumes linearity,
which means they can be added and scale them.

The manifold is the local vector space of a set. While a vector space assumes linearity, a manifold
can be non-linear but resembles the local vector space. 

In PAC, we scaled the data and placed it on a vector space by rotating individual points.
In T-SNE, it assumes the data lies on a manifold and preserves local neighborhood 
relationships to calculate the probability
that two points are neighbors.

## Question 4
#### a) Perform T-SNE
Set `seed = 123`.  
*Hint:* Subset your PCA results to PC1–PC10, add the class variable back in, remove duplicates, then perform T-SNE.

```{r}
library(Rtsne)

wine_PCA_DF <- PCA_wine$x %>% as.data.frame() %>% select(PC1:PC10)

#adding wine class back
wine_PCA_DF$class <- wine$class

#removing duplicates
wine_PCA_no_duplicates <- wine_PCA_DF %>% filter(!duplicated(.))

str(wine_PCA_no_duplicates)

set.seed(123)

tsne_wine <- wine_PCA_no_duplicates %>% select(-class) %>% 
                  Rtsne(.,
                  dims = 2,
                  perplexity = 30,
                  verbose = FALSE,
                  check_duplicates = FALSE
                )


```


#### b) Plot the results in 2D
*Hint:* Convert your T-SNE results to a tibble and add back the class variable from your scaled data set using a `mutate()` statement. Then, you can use `color = factor(class)` in your `ggplot` statement.

```{r}
results <- as_tibble(tsne_wine$Y)

ggplot(results, aes(V1, V2)) + geom_point(aes(color = factor(wine_PCA_DF$class)))
```


#### c) Why didn't we stop at PCA?
We didn't stop at PCA because while PCA is great at reducing redundancy and dimensionality in the data, 
T-SNE is better are visualizing the complex spacial relationship of this data by
predicting the probability of two points being neighbors. This helps visualize
the similarities and differences of the wine classes in the original data.

#### d) What other types of data does this workflow make sense for?
Genomics and gene expression data would make sense
because of the complex biological relationships/clusters that exist.
Another type of data would be neural activity data because it has
high dimensionality and noisiness but could be reduced to determine clustering.


