---
title: BIOS Homework 10
output: pdf_document
---

# Homework 10

## Student: Jama Brookes

#### Course Notes
**Language Models:** https://github.com/rjenki/BIOS512/tree/main/lecture17  
**Unix:** https://github.com/rjenki/BIOS512/tree/main/lecture18  
**Docker:** https://github.com/rjenki/BIOS512/tree/main/lecture19

## Question 1
#### Make a language model that uses ngrams and allows the user to specify start words, but uses a random start if one is not specified.
```{r}
# required libraries
library(httr)
library(tokenizers)

```


#### a) Make a function to tokenize the text.
```{r}
tokenize_text <- function(text) {
    tokenizers::tokenize_words(text, lowercase=TRUE, strip_punct=TRUE)[[1]]
}

```


#### b) Make a function generate keys for ngrams.
```{r}
key_from <- function(ngram, sep = "\x1f") {
    paste(ngram, collapse=sep)
}
```


#### c) Make a function to build an ngram table.
```{r}
build_ngram_table <- function(tokens, n, sep = "\x1f") {
    if (length(tokens) < n) return(new.env(parent = emptyenv()))
    tbl <- new.env(parent = emptyenv())
    for (i in seq_len(length(tokens) - n + 1L)) {
        ngram <- tokens[i:(i + n - 2L)]
        next_word <- tokens[i + n - 1L]
        key <- paste(ngram, collapse = sep)
        counts <- if (!is.null(tbl[[key]])) tbl[[key]] else integer(0)
        if (next_word %in% names(counts)) {
            counts[[next_word]] <- counts[[next_word]] + 1L
        } else {
            counts[[next_word]] <- 1L
        }
        tbl[[key]] <- counts
    }
    tbl
}
```


#### d) Function to digest the text.
```{r}
digest_text <- function(text, n) {
    tokens <- tokenize_text(text)
    build_ngram_table(tokens, n)
}
```


#### e) Function to digest the url.
```{r}
digest_url <- function(url, n) {
    res <- httr::GET(url)
    txt <- httr::content(res, as = "text", encoding = "UTF-8")
    digest_text(txt,n)
}
```


#### f) Function that gives random start.
```{r}
random_start <- function(tbl, sep = "\x1f") {
    keys <- ls(envir = tbl, all.names=TRUE)
    if (length(keys)==0) stop("No n-grams available. Digest text first.")
    picked <- sample(keys, 1)
    strsplit(picked, sep, fixed=TRUE)[[1]]
}
```


#### g) Function to predict the next word.
```{r}
predict_next_word <- function(tbl, ngram, sep = "\x1f") {
    key <- paste(ngram, collapse = sep)
    counts <- if(!is.null(tbl[[key]])) tbl[[key]] else integer(0)
    if (length(counts) == 0) return(NA_character_)
    sample(names(counts), size=1, prob=as.numeric(counts))
}
```

#### h) Function that puts everything together. Specify that if the user does not give a start word, then the random start will be used.
```{r}
make_ngram_generator <- function(tbl, n, sep = "\x1f") {
    force(tbl); n <- as.integer(n); force(sep)
    function(start_words = NULL, length = 10L) {
        if ((is.null(start_words)) || length(start_words) != n - 1L) {
            start_words <- random_start(tbl, sep=sep)
        }
        word_sequence <- start_words
        for (i in seq_len(max(0L, length - length(start_words)))) {
            ngram <- tail(word_sequence, n - 1L)
            next_word <- predict_next_word(tbl, ngram, sep=sep)
            if (is.na(next_word)) break
            word_sequence <- c(word_sequence, next_word)
        }
        paste(word_sequence, collapse= " ")
    }
}
```


## Question 2
#### For this question, set `seed=2025`.
#### a) Test your model using a text file of [Grimm's Fairy Tails](https://www.gutenberg.org/cache/epub/2591/pg2591.txt)
```{r}
set.seed(2025)
url <- "https://www.gutenberg.org/cache/epub/2591/pg2591.txt" # Grimm's Fairy Tails
tbl3 <- digest_url(url, n=3)
gen3 <- make_ngram_generator(tbl3, n=3)
```

#### i) Using n=3, with the start word(s) "the king", with length=15. 
```{r}
output <- gen3(start_words = c("the", "king"), length = 15)
cat(output)
```

#### ii) Using n=3, with no start word, with length=15.
```{r}
output <- gen3(length = 15)
cat(output)

```


#### b) Test your model using a text file of [Ancient Armour and Weapons in Europe](https://www.gutenberg.org/cache/epub/46342/pg46342.txt)
```{r}
set.seed(2025)
url <- "https://www.gutenberg.org/cache/epub/46342/pg46342.txt" #Ancient Armour & Weapons in Europe
tbl3 <- digest_url(url, n=3)
gen4 <- make_ngram_generator(tbl3, n=3)
```

#### i) Using n=3, with the start word(s) "the king", with length=15. 
```{r}
set.seed(2025)
output <- gen4(start_words = c("the", "king"), length = 15)
cat(output)
```

#### ii) Using n=3, with no start word, with length=15.
```{r}
set.seed(2025)
output <- gen4(length = 15)
cat(output)
```

#### c) Explain in 1-2 sentences the difference in content generated from each source.

In the Grimm's Fairy Tails, the content has more of a ficitonal plot from a story, which is whimisical in nature. The Ancient Armour & Weapons in Europe produces more a more direct story that likely it talking about historical events and weaponry, like swords.

## Question 3
#### a) What is a language learning model? 
A language model simply predicts the next word by using a probability distribution.

#### b) Imagine the internet goes down and you can't run to your favorite language model for help. How do you run one locally?
We could run a language model locally by training one on information from public domain (like the ones on the internet today). We could do this by downloading OLLAMA, which uses language model tools, and then train it ourselves.

## Question 4
#### Explain what the following vocab words mean in the context of typing `mkdir project` into the command line. If the term doesn't apply to this command, give the definition and/or an example.
| Term | Meaning |  
|------|---------|
| **Shell** | The shell takes the input `mkdir project` and parses out what system to run/use. |
| **Terminal emulator** | Is the actual interface that you type mkdir project into. |
| **Process** | actually runs the code by creating a new "process" or running program. |
| **Signal** | we send these to tell the program to do something. |
| **Standard input** | reads the characters from the input. |
| **Standard output** | outputs characters if needed (typically no output is produce with mkdir commands) |
| **Command line argument** | is the thing we add to tell the process what to do. For the example, `project` |
| **The environment** | all the stuff a process can see when its running. |

## Question 5
#### Consider the following command `find . -iname "*.R" | xargs grep read_csv`.
#### a) What are the programs?
find, xargs, and grep

#### b) Explain what this command is doing, part by part.
`find .` -  searches for the file in the current directory 
`-iname` - (with name of) ".R" (aka R files)
`|` - pipe
`xargs` - takes the files found
`grep read_csv` - searchs for texts in the files and reads them


## Question 6
#### Install Docker on your machine. See [here](https://github.com/rjenki/BIOS512/blob/main/lecture18/docker_install.md) for instructions. 
#### a) Show the response when you run `docker run hello-world`.
Hello from Docker!
This message shows that your installation appears to be working correctly.

To generate this message, Docker took the following steps:
 1. The Docker client contacted the Docker daemon.
 2. The Docker daemon pulled the "hello-world" image from Docker Hub.
 3. The Docker daemon created a new container from that image.
 4. The Docker daemon streamed that output to the Docker client.
 
 Server is ready

#### b) Access Rstudio through a Docker container. Set your password and make sure your files show up on the Rstudio server. Type the command and the output you get below.
docker run --platform linux/amd64 -it -e PASSWORD=dockertrial -p 8787:8787 rocker/verse

Adding password for user rstudio
useradd: warning: the home directory already exists.
Not copying any file from skel directory into it.
rsession: no process found

### RStudio Server Info ###
Version: 2024.04.0+492
Address: 0.0.0.0
Port: 8787
Username: rstudio
Password: dockertrial

RStudio Server is ready to accept connections
― open your browser and go to http://localhost:8787


#### c) How do you log in to the RStudio server?
I went to the local host link and logged in with the username and password above.

